---
title: "GEB main analysis"
subtitle: "Key results and workflow checks"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

This report presents selected results from the completed workflow. Full tables
and publication figures are exported separately under `outputs/`.

```{r}
library(dplyr)

lcbd_results_path <- targets::tar_read(lcbd_results_file)
regional_results_path <- targets::tar_read(regional_results_file)
gllvm_metadata_path <- targets::tar_read(gllvm_metadata_file)
publication_paths <- targets::tar_read(publication_files)

lcbd_results <- readRDS(lcbd_results_path)
regional_results <- readRDS(regional_results_path)
gllvm_metadata <- readRDS(gllvm_metadata_path)

component_labels <- c(
  total = "Total beta diversity",
  replacement = "Replacement",
  richness_difference = "Richness difference"
)

publication_figure <- function(file_name, alt_text) {
  matched_path <- publication_paths[basename(publication_paths) == file_name]

  if (length(matched_path) != 1L) {
    stop("Expected one publication file named: ", file_name, call. = FALSE)
  }

  # targets stores paths relative to the project root; Pandoc renders this
  # document from reports/, so its image links need one parent-directory step.
  report_path <- file.path("..", unname(matched_path))
  report_path <- gsub("\\\\", "/", report_path)
  knitr::asis_output(sprintf("![%s](%s)", alt_text, report_path))
}
```

## Analysis overview

The tables below report the dimensions recorded in each completed analysis
checkpoint. They provide a quick check of the site, species, model and allocation
counts used in the final analyses.

### LCBD dimensions

```{r}
knitr::kable(lcbd_results$dimensions)
```

### Regional convergence dimensions

```{r}
knitr::kable(regional_results$dimensions)
```

### GLLVM dimensions

```{r}
knitr::kable(gllvm_metadata$dimensions)
```

## Local contributions to beta diversity

The LCBD models compare pool-size-standardised local uniqueness across four
focal gradients. The interval combines uncertainty within models and variation
among the repeated spatial reference-pool allocations.

```{r}
lcbd_table <- lcbd_results$main_results |>
  mutate(
    lcbd_component = unname(component_labels[lcbd_component])
  ) |>
  select(
    focal_gradient = focal_label,
    lcbd_component,
    n_allocations,
    n_unique_partitions,
    mean_n_sites,
    mean_n_informative_pools,
    mean_endpoint_difference,
    lower_95,
    upper_95,
    direction_agreement,
    interpretation
  ) |>
  arrange(focal_gradient, lcbd_component)

knitr::kable(lcbd_table, digits = 3)
```

```{r}
#| results: asis
publication_figure("04_local_uniqueness.svg", "LCBD response curves")
```

### Reference-pool checks

Negative component LCBD values are retained as diagnostics. The enforced
numerical check is that raw LCBD sums to one
within every reference pool and component.

```{r}
lcbd_pool_checks <- lcbd_results$pool_lcbd_checks |>
  group_by(lcbd_component) |>
  summarise(
    n_pool_components = n(),
    maximum_absolute_sum_error = max(abs(raw_lcbd_sum - 1)),
    minimum_raw_lcbd = min(minimum_raw_lcbd),
    n_values_below_negative_tolerance =
      sum(n_values_below_negative_tolerance),
    .groups = "drop"
  ) |>
  mutate(
    lcbd_component = unname(component_labels[lcbd_component])
  )

knitr::kable(lcbd_pool_checks, digits = 6)
```

## Regional convergence

The regional analysis compares beta diversity among IBRA7 subregions at the low
and high endpoints of each focal gradient.

```{r}
regional_table <- regional_results$endpoint_contrasts |>
  select(
    gradient = gradient_label,
    component,
    low_end_mean,
    high_end_mean,
    high_minus_low,
    difference_lower_95,
    difference_upper_95,
    bootstrap_two_sided_p_value,
    interpretation
  )

knitr::kable(regional_table, digits = 3)
```

```{r}
#| results: asis
publication_figure("05_regional_convergence_2x2.svg", "Regional convergence")
```

## GLLVM/JSDM

The final model checkpoint records the retained species, occurrence support,
trait coverage, environmental predictors and selected spatial variables. Full
coefficient tables and publication figures are produced by the dedicated GLLVM
output stage.

```{r}
gllvm_species_summary <- gllvm_metadata$species_counts |>
  summarise(
    n_species = n(),
    minimum_presences = min(n_presence),
    median_presences = median(n_presence),
    maximum_presences = max(n_presence),
    minimum_occupancy = min(occupancy),
    median_occupancy = median(occupancy),
    maximum_occupancy = max(occupancy)
  )

knitr::kable(gllvm_species_summary, digits = 3)
```

```{r}
trait_matching <- data.frame(
  measure = c(
    "Species retained in both GLLVM models",
    "Species excluded because traits were unavailable"
  ),
  value = c(
    length(gllvm_metadata$species_order),
    length(gllvm_metadata$missing_traits)
  )
)

knitr::kable(trait_matching)
```

### Species responses

```{r}
#| results: asis
publication_figure("beta_heatmap_selected_grouped.svg", "Species responses")
```

### Trait responses

```{r}
#| results: asis
publication_figure("trait_heatmap_selected.svg", "Trait responses")
```

## Reproducibility information

```{r}
reproducibility <- data.frame(
  setting = c(
    "LCBD allocations",
    "Regional bootstrap draws",
    "GLLVM occupancy threshold"
  ),
  value = c(
    lcbd_results$settings$n_reference_pool_allocations,
    regional_results$settings$bootstrap_draws,
    gllvm_metadata$occupancy_threshold
  )
)

knitr::kable(reproducibility)
```
